Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-13 tsibbletalk.Rmd
more bibs

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address:
  - Department of Statistics
  - The University of Auckland
  - New Zealand
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address:
  - Department of Econometrics and Business Statistics
  - Monash University
  - Australia
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  An abstract of less than 150 words.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse} [@Wickham2019], the **tidyverts** suite features `tsibble` as the foundational data structure, in order to build a fluid and fluent pipeline for time series analysis. Besides \CRANpkg{tsibble}, the \CRANpkg{feasts} and \CRANpkg{fable} packages fill the role of statistical analysis and forecasting in the **tidyverts** ecosystem. When time series analysis starts taking off, series of interest denoted by the `key` variables often remain unchanged over the course of analysis, from trend inspection to forecasting performance.

@@

R \CRANpkg{shiny} and \CRANpkg{htmlwidgets} lay a fundamental infrastructure in marrying R with HTML elements and Javascript for interactivity, although they do not emphasise on interactive graphics. The \CRANpkg{htmlwidgets} package opens a door to embedding Javascript libraries into R such that users are able to write only R code to generate web-based plots. Many Javascript charting libraries have been ported to R as HTML widgets, including \CRANpkg{plotly}, \CRANpkg{rbokeh}, and \CRANpkg{leaflet} for maps. To enable interactions between different widgets, it can be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk}. The \CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6 instances to support linked brushing and filtering across widgets.

If not using web technology, two examples are provided here: \CRANpkg{loon} based on Tcl/Tk and \CRANpkg{cranvas} based on Qt. They offer a wide array of pre-defined interactions, such as selecting and zooming, to manipulate plots via mouse, keyboard strokes, and menus. The \CRANpkg{cranvastime} is an add-on of \CRANpkg{cranvas}, which develops a set of novel interactions for temporal data, for example wrapping and mirroring.

@@

The `tourism_monthly` dataset packaged in \CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across Australia, to give an illustrator of nesting and crossing. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

@@

The city of Melbourne has sensors installed to count hourly tallies of pedestrians in order to capture downtown daily rhythms (cite). Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four locations, with the depiction of three pronounced slices in time. Figure \ref{fig:wrap-ped-1} unfolds all counts from January to May on their absolute time line, facetted by four sensors. On March 16, Melbourne went to the stage three lockdown due to COVID-19, seeing a significant decline in traffic volume across the city. These lines are then folded into daily and weekly sections, shown as Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal variations have been popped out to viewers, complementing the not-just-magnitude-drop story. The pre-lockdown period is coloured with dark green and lockdown with orange.

@@

This shiny module, decoupled to `tsibbleDiceUI()` and `tsibbleDiceServer()`, presents a clean interface and forms a resusable piece in a shiny application. A shiny module provides the vehicle in modularising shiny applications for both users and developers. Like all shiny modules, the first argument in both functions requires a user-supplied session id that must be unique. The UI function `tsibbleDiceUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is certainly the server function `tsibbleDiceServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, and thereby the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. As for date-times (represented by `POSIXt`), it ranges from from fine "day" to coarse "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

>>>>

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Auckland, New Zealand
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Melbourne, Australia
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  An abstract of less than 150 words.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package [@R-tsibbledata]. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse} [@Wickham2019], the **tidyverts** suite features `tsibble` as the foundational data structure, in order to build a fluid and fluent pipeline for time series analysis. Besides \CRANpkg{tsibble}, the \CRANpkg{feasts} [@R-feasts] and \CRANpkg{fable} [@R-fable] packages fill the role of statistical analysis and forecasting in the **tidyverts** ecosystem. When time series analysis starts taking off, series of interest denoted by the `key` variables often remain unchanged over the course of analysis, from trend inspection to forecasting performance.

@@

R \CRANpkg{shiny} [@R-shiny] and \CRANpkg{htmlwidgets} [@R-htmlwidgets] lay a fundamental infrastructure in marrying R with HTML elements and Javascript for interactivity, although they do not emphasise on interactive graphics. The \CRANpkg{htmlwidgets} package opens a door to embedding Javascript libraries into R such that users are able to write only R code to generate web-based plots. Many Javascript charting libraries have been ported to R as HTML widgets, including \CRANpkg{plotly} [@plotly2020], \CRANpkg{rbokeh} [@R-rbokeh], and \CRANpkg{leaflet} [@R-leaflet] for maps. To enable interactions between different widgets, it can be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk} [@R-crosstalk]. The \CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6 instances to support linked brushing and filtering across widgets, without relying on \CRANpkg{shiny}.

If not using web technology, two examples are provided here: \CRANpkg{loon} [@R-loon] based on Tcl/Tk and \pkg{cranvas} [@xie_reactive_2014] based on Qt. They offer a wide array of pre-defined interactions, such as selecting and zooming, to manipulate plots via mouse, keyboard strokes, and menus. The \pkg{cranvastime} package [@cheng_enabling_2016] is an add-on of \pkg{cranvas}, which develops a set of novel interactions for temporal data, for example wrapping and mirroring.

@@

The `tourism_monthly` dataset [@tourism] packaged in \CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across Australia, to give an illustrator of nesting and crossing. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

@@

The city of Melbourne has sensors installed to count hourly tallies of pedestrians in order to capture downtown daily rhythms [@ped]. Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four locations, with the depiction of three pronounced slices in time. Figure \ref{fig:wrap-ped-1} unfolds all counts from January to May on their absolute time line, facetted by four sensors. On March 16, Melbourne went to the stage three lockdown due to COVID-19, seeing a significant decline in traffic volume across the city. These lines are then folded into daily and weekly sections, shown as Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal variations have been popped out to viewers, complementing the not-just-magnitude-drop story. The pre-lockdown period is coloured with dark green and lockdown with orange.

@@

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a resusable piece in a shiny application. A shiny module provides the vehicle in modularising shiny applications for both users and developers. Like all shiny modules, the first argument in both functions requires a user-supplied session id that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is certainly the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, and thereby the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. As for date-times (represented by `POSIXt`), it ranges from from fine "day" to coarse "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

```{r pkg-bib, eval = FALSE}
pkgs <- c("tsibbledata", "feasts", "fable", "shiny", "htmlwidgets", "plotly",
  "rbokeh", "leaflet", "crosstalk", "loon")
write_bib(pkgs, "rpkgs.bib")
```

### 2020-09-13 tsibbletalk.tex
more bibs

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package.
The \texttt{Month} column holds year-months as \texttt{index}. The
\texttt{State} together with \texttt{Industry} are the identifiers for
these 152 series, highlighted as \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, in order to build a
fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} and \CRANpkg{fable} packages
fill the role of statistical analysis and forecasting in the
\textbf{tidyverts} ecosystem. When time series analysis starts taking
off, series of interest denoted by the \texttt{key} variables often
remain unchanged over the course of analysis, from trend inspection to
forecasting performance.

@@

R \CRANpkg{shiny} and \CRANpkg{htmlwidgets} lay a fundamental
infrastructure in marrying R with HTML elements and Javascript for
interactivity, although they do not emphasise on interactive graphics.
The \CRANpkg{htmlwidgets} package opens a door to embedding Javascript
libraries into R such that users are able to write only R code to
generate web-based plots. Many Javascript charting libraries have been
ported to R as HTML widgets, including \CRANpkg{plotly},
\CRANpkg{rbokeh}, and \CRANpkg{leaflet} for maps. To enable interactions
between different widgets, it can be achieved with \CRANpkg{shiny} or
\CRANpkg{crosstalk}. The \CRANpkg{crosstalk} extends
\CRANpkg{htmlwidgets} with shared R6 instances to support linked
brushing and filtering across widgets.

If not using web technology, two examples are provided here:
\CRANpkg{loon} based on Tcl/Tk and \CRANpkg{cranvas} based on Qt. They
offer a wide array of pre-defined interactions, such as selecting and
zooming, to manipulate plots via mouse, keyboard strokes, and menus. The
\CRANpkg{cranvastime} is an add-on of \CRANpkg{cranvas}, which develops
a set of novel interactions for temporal data, for example wrapping and
mirroring.

@@

The \texttt{tourism\_monthly} dataset packaged in \CRANpkg{tsibbletalk},
contains monthly domestic overnight trips across Australia, to give an
illustrator of nesting and crossing. The \texttt{key} is comprised of
three identifying variables: \texttt{State}, \texttt{Region}, and
\texttt{Purpose} (of trip), in particular \texttt{State} nesting of
\texttt{Region}, together crossed with \texttt{Purpose}. This
specification can be translated as follows:

@@

The city of Melbourne has sensors installed to count hourly tallies of
pedestrians in order to capture downtown daily rhythms (cite). Figure
\ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at
four locations, with the depiction of three pronounced slices in time.
Figure \ref{fig:wrap-ped-1} unfolds all counts from January to May on
their absolute time line, facetted by four sensors. On March 16,
Melbourne went to the stage three lockdown due to COVID-19, seeing a
significant decline in traffic volume across the city. These lines are
then folded into daily and weekly sections, shown as Figure
\ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal
variations have been popped out to viewers, complementing the
not-just-magnitude-drop story. The pre-lockdown period is coloured with
dark green and lockdown with orange.

@@

This shiny module, decoupled to \texttt{tsibbleDiceUI()} and
\texttt{tsibbleDiceServer()}, presents a clean interface and forms a
resusable piece in a shiny application. A shiny module provides the
vehicle in modularising shiny applications for both users and
developers. Like all shiny modules, the first argument in both functions
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleDiceUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is certainly the server
function \texttt{tsibbleDiceServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, and thereby the time
\texttt{index} can be retrieved for dissection. The \texttt{period}
option semantically takes a desired number of seasonal periods to be
shifted, for example data shifted by ``1 day'', ``2 days'', or ``1
week'', etc. In other words, the \texttt{period} defines the grind
level. As for date-times (represented by \texttt{POSIXt}), it ranges
from from fine ``day'' to coarse ``year''. The following code snippet
generates Figure \ref{fig:wrap-ped}.

@@

\address{%
Earo Wang\\
The University of Auckland\\%
Department of Statistics\\ The University of Auckland\\ New Zealand\\
%
%
%
\\\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\ Monash
University\\ Australia\\
%
%
%
\\\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
}

>>>>

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} classes with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package
\citep{R-tsibbledata}. The \texttt{Month} column holds year-months as
\texttt{index}. The \texttt{State} together with \texttt{Industry} are
the identifiers for these 152 series, highlighted as \texttt{key}. Note
that the column \texttt{Series\ ID} could be an alternative option for
setting up \texttt{key}, but \texttt{State} and \texttt{Industry} are
more readable and informative. The \texttt{index} and \texttt{key} are
``sticky'' columns to a \texttt{tsibble}, forming critical pieces for
fluent temporal data analysis later.

@@

In the spirit of tidy data to the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, in order to build a
fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} \citep{R-feasts} and
\CRANpkg{fable} \citep{R-fable} packages fill the role of statistical
analysis and forecasting in the \textbf{tidyverts} ecosystem. When time
series analysis starts taking off, series of interest denoted by the
\texttt{key} variables often remain unchanged over the course of
analysis, from trend inspection to forecasting performance.

@@

R \CRANpkg{shiny} \citep{R-shiny} and \CRANpkg{htmlwidgets}
\citep{R-htmlwidgets} lay a fundamental infrastructure in marrying R
with HTML elements and Javascript for interactivity, although they do
not emphasise on interactive graphics. The \CRANpkg{htmlwidgets} package
opens a door to embedding Javascript libraries into R such that users
are able to write only R code to generate web-based plots. Many
Javascript charting libraries have been ported to R as HTML widgets,
including \CRANpkg{plotly} \citep{plotly2020}, \CRANpkg{rbokeh}
\citep{R-rbokeh}, and \CRANpkg{leaflet} \citep{R-leaflet} for maps. To
enable interactions between different widgets, it can be achieved with
\CRANpkg{shiny} or \CRANpkg{crosstalk} \citep{R-crosstalk}. The
\CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6
instances to support linked brushing and filtering across widgets,
without relying on \CRANpkg{shiny}.

If not using web technology, two examples are provided here:
\CRANpkg{loon} \citep{R-loon} based on Tcl/Tk and \pkg{cranvas}
\citep{xie_reactive_2014} based on Qt. They offer a wide array of
pre-defined interactions, such as selecting and zooming, to manipulate
plots via mouse, keyboard strokes, and menus. The \pkg{cranvastime}
package \citep{cheng_enabling_2016} is an add-on of \pkg{cranvas}, which
develops a set of novel interactions for temporal data, for example
wrapping and mirroring.

@@

The \texttt{tourism\_monthly} dataset \citep{tourism} packaged in
\CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across
Australia, to give an illustrator of nesting and crossing. The
\texttt{key} is comprised of three identifying variables:
\texttt{State}, \texttt{Region}, and \texttt{Purpose} (of trip), in
particular \texttt{State} nesting of \texttt{Region}, together crossed
with \texttt{Purpose}. This specification can be translated as follows:

@@

The city of Melbourne has sensors installed to count hourly tallies of
pedestrians in order to capture downtown daily rhythms \citep{ped}.
Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot
traffic at four locations, with the depiction of three pronounced slices
in time. Figure \ref{fig:wrap-ped-1} unfolds all counts from January to
May on their absolute time line, facetted by four sensors. On March 16,
Melbourne went to the stage three lockdown due to COVID-19, seeing a
significant decline in traffic volume across the city. These lines are
then folded into daily and weekly sections, shown as Figure
\ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal
variations have been popped out to viewers, complementing the
not-just-magnitude-drop story. The pre-lockdown period is coloured with
dark green and lockdown with orange.

@@

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
resusable piece in a shiny application. A shiny module provides the
vehicle in modularising shiny applications for both users and
developers. Like all shiny modules, the first argument in both functions
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is certainly the server
function \texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, and thereby the time
\texttt{index} can be retrieved for dissection. The \texttt{period}
option semantically takes a desired number of seasonal periods to be
shifted, for example data shifted by ``1 day'', ``2 days'', or ``1
week'', etc. In other words, the \texttt{period} defines the grind
level. As for date-times (represented by \texttt{POSIXt}), it ranges
from from fine ``day'' to coarse ``year''. The following code snippet
generates Figure \ref{fig:wrap-ped}.

@@

\address{%
Earo Wang\\
The University of Auckland\\%
Auckland, New Zealand\\
%
%
%
\\\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}
}

\address{%
Dianne Cook\\
Monash University\\%
Melbourne, Australia\\
%
%
%
\\\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
}
